---
output: github_document
---

<!-- README.md is generated from README.Rmd — edit the .Rmd and run `make readme`.
     SQL blocks below are executed live against the built extension via duckknit. -->

> [!IMPORTANT]
> **This repository is archived.** `duckvep` has moved from this Rust prototype to a
> C / DuckDB-extension implementation built on
> [`duckhts`](https://github.com/RGenomicsETL/duckhts) (htslib-backed) plus the
> `duckhgvs` engine. The active successor is
> **➡️ [RGenomicsETL/duckvep](https://github.com/RGenomicsETL/duckvep)**.
> This repo remains as design reference and as a differential-test **oracle**
> (its consequence engine + sweep prototypes are run to diff the C kernel against).

```{r setup, include=FALSE}
options(duckknit.duckdb = normalizePath("scripts/duckdb-unsigned"))
knitr::knit_engines$set(sql = duckknit::eng_duckdb)
knitr::opts_chunk$set(mode = "markdown", session = "readme",
                      comment = "", results = "asis")
```

```{r conformance-numbers, include=FALSE}
# Headline conformance figures are READ from the recorded reports, never typed in the prose —
# so they can't go stale when the harness is re-run (correctness/vep_concordance.R writes them).
.ma  <- read.csv("correctness/data/methodology_audit.csv", stringsAsFactors = FALSE)
.mav <- setNames(.ma$value, .ma$metric)
.ci  <- read.csv("correctness/data/concordance_by_impact.csv", stringsAsFactors = FALSE)
n_variants  <- format(max(.ci$n_variants), big.mark = ",")
total_pairs <- sprintf("%.2fM", sum(.ci$pairs[.ci$engine == "duckvep"]) / 1e6)
k <- function(m) format(.mav[[m]], big.mark = ",")
dv_total <- k("duckvep_total_divergence"); dv_shared <- k("duckvep_discordant_on_shared")
dv_emit  <- k("duckvep_only_pairs_emission"); dv_spec  <- k("duckvep_specific_on_shared")
fv_total <- k("fastvep_total_divergence")
```

# duckvep — VEP-style variant consequences in DuckDB

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

duckvep is a loadable DuckDB extension, written in Rust, that implements Ensembl-VEP-style
consequence prediction and HGVS as SQL functions. It reads genomics formats with
[noodles](https://github.com/zaeleus/noodles) and treats annotation databases as ordinary
Parquet/DuckDB tables that the query optimizer joins. The consequence engine is a Rust port of
[fastVEP](https://github.com/Huang-lab/fastVEP) with accuracy patches documented in
[docs/PATCHES.md](docs/PATCHES.md). It is under active development; local builds require DuckDB
v1.5.3 with unsigned extension loading.

Concordance with Ensembl VEP is measured rather than assumed. On a controlled `--gff` setup, where
VEP, duckvep and the vendored fastVEP read the same gene model and results are keyed to the original
input variant, duckvep diverges from VEP on `r dv_total` of roughly `r total_pairs` variant/transcript
pairs (N=`r n_variants` ClinVar variants): `r dv_shared` with a different consequence on a shared pair
plus `r dv_emit` emission differences. Of the shared-pair disagreements, `r dv_spec` are
duckvep-specific (fastVEP matches VEP where duckvep does not); fastVEP itself diverges on `r fv_total`
pairs. The remaining duckvep cases are boundary indels where VEP 3'-shifts the allele before calling
consequence. The reports under [`conformance/`](conformance/) are stratified by consequence term,
variant type and length bin with exact 95% confidence intervals; the framework is a differential
falsifier and an empirical bound, not a proof. Known gaps are tracked in
[`correctness/`](correctness/correctness.md) and [`conformance/`](conformance/).

Beyond single-variant calls, `vep_haplotype_consequence` applies phased variants together, in the
style of bcftools-`csq` and haplosaurus, so a silent SNV becomes missense when phased with its
neighbour; this path is experimental. Annotation sources — gnomAD, ClinVar, conservation, a custom
BED file — are Parquet or DuckDB tables joined on `normalize_variant`'s key and composed in SQL
through a `sources.sql` manifest and an `annotate_vcf()` macro, rather than a command-line flag per
source, and duckvep can share a DuckDB session with the duckhts extension. The transcript cache is
built from Ensembl MySQL dumps and inherits the curated MANE, `cds_start_NF` and selenocysteine
flags. The implementation is pure Rust; a WASM target exists but is experimental and not yet covered
by CI.

See [docs/DESIGN.md](docs/DESIGN.md), [NEWS.md](NEWS.md), [docs/PATCHES.md](docs/PATCHES.md) and
[correctness/](correctness/correctness.md).

## Build

```sh
make debug      # builds build/debug/duckvep.duckdb_extension (native)
make test       # runs the SQL test suite
```

A WASM target exists, but browser/DuckDB-WASM packaging is experimental and not yet
covered by CI.

## Load it

```{sql}
LOAD 'build/debug/duckvep.duckdb_extension';
```

## `read_vcf` — VCF/BCF as a SQL table

One row per variant. `alt` and `filter` are lists; `end_pos` carries the variant
interval (`INFO/END` for SV/CNV, else `pos + len(ref) - 1`).

```{sql}
SELECT chrom, pos, end_pos, ref, alt, qual, filter
FROM read_vcf('test/data/sites.vcf')
LIMIT 5;
```

### Structural variants

Symbolic (`<DEL>`, `<CNV>`), breakend, and multiallelic alleles survive as list
elements, and `end_pos` makes interval filters work:

```{sql}
SELECT pos, end_pos, ref, alt, end_pos - pos AS span
FROM read_vcf('test/data/sv.vcf')
WHERE end_pos - pos > 100
ORDER BY pos;
```

### Multi-sample, phased genotypes

`gt` is the positional per-sample genotype list (phasing preserved as `0|1`):

```{sql}
SELECT pos, ref, alt, gt
FROM read_vcf('test/data/ms.vcf');
```

`vcf_samples()` gives the header sample order, so genotypes explode to tidy
per-sample rows on demand — no re-annotation, because annotation is site-wise:

```{sql}
SELECT v.pos, s.sample, g.gt
FROM read_vcf('test/data/ms.vcf') v,
     UNNEST(v.gt) WITH ORDINALITY AS g(gt, idx)
     JOIN vcf_samples('test/data/ms.vcf') s USING (idx)
ORDER BY v.pos, s.idx;
```

### Region filter

```{sql}
SELECT count(*) AS n
FROM read_vcf('test/data/sv.vcf', region := 'chr1:4000-13000');
```

## Variant effect prediction

`vep_load_cache('<gene model>', '<fasta>')` builds the consequence engine **once**
into the connection's state, so later `vep_consequence`/`vep_annotate` calls reuse
it (the result `loaded` is just a confirmation). The gene model is either a GFF3
or duckvep's **columnar Parquet cache**: a GFF3 is parsed *and* written to
`<gff3>.transcripts.parquet` for next time; pass that `.parquet` to skip the parse.
The FASTA (`''` = none) is needed for exact coding calls (synonymous vs missense).

```{sql}
-- first call parses test.gff3 and writes test.gff3.transcripts.parquet;
-- next time, load the cache directly: vep_load_cache('test.gff3.transcripts.parquet','')
SELECT vep_load_cache('test/data/test.gff3', '');
```

Then `vep_consequence(chrom, pos, ref, alt)` is a scan-driven scalar returning a
native `LIST<STRUCT>` — `UNNEST` it. Driven by DuckDB's scan, so variants come
from `read_vcf`, `read_parquet`, or any relation.

Each struct also carries HGVS notation: `hgvsg` (genomic), `hgvsc` (coding) and
`hgvsp` (protein, when a FASTA is loaded). On the chr17 benchmark dataset the HGVSc
and HGVSp output matches fastVEP for all recorded comparisons (see
[`benchmarks/results.md`](benchmarks/results.md)):

```{sql}
SELECT c.transcript_id, c.consequence, c.impact, c.hgvsg, c.hgvsc
FROM UNNEST(vep_consequence('17', 43124090, 'A', 'G')) AS u(c)
WHERE c.gene_symbol = 'BRCA1'
ORDER BY c.canonical DESC
LIMIT 4;
```

Or annotate a whole VCF in one call with `vep_annotate(vcf, gff3 := …)`, one row
per (variant, transcript, allele). Concordance against Ensembl VEP is reported below:

```{sql}
SELECT pos, ref, alt, gene_symbol, transcript_id, consequence, impact
FROM vep_annotate('test/data/sites.vcf', gff3 := 'test/data/test.gff3')
WHERE canonical AND impact <> 'MODIFIER'
ORDER BY pos LIMIT 5;
```

### Structural variants

Symbolic SVs (`<DEL>`, `<DUP>`, `<CNV>`, `<INV>`) and breakends are dispatched to
the SV consequence predictor over their full `INFO/END` span — `vep_annotate`
reads `END` from the VCF, and the scalar takes an END-aware 5-argument form
`vep_consequence(chrom, pos, end, ref, alt)` (drive it with `read_vcf`'s
`end_pos`). A `<DEL>` covering a transcript yields `transcript_ablation`, a
partial one `feature_truncation` + `splice_*` + `copy_number_decrease`, and a
`<DUP>` `transcript_amplification` — the Ensembl SV vocabulary:

```{sql}
SELECT c.consequence, c.impact
FROM UNNEST(vep_consequence('17', 43044295, 43125483, 'N', '<DEL>')) AS u(c)
WHERE c.transcript_id = 'ENST00000357654';
```

## Variants from any source

Because the variant table is just columns, anything DuckDB can read is a valid
variant provider — `read_vcf`, `read_parquet`, `read_csv`, an attached DB, or a
literal relation. The downstream UDFs consume the columns, not the format:

```{sql}
SELECT chrom, pos, ref, alt
FROM (VALUES ('chr1', 100, 'A', ['G']),
             ('chr2', 200, 'C', ['T', 'TA'])) AS t(chrom, pos, ref, alt);
```

## Composing with duckhts

duckvep can be used with **[duckhts](https://github.com/sounkou-bioinfo/duckhts)**
(htslib-backed format readers, a DuckDB community extension) in the same DuckDB
session: load both extensions and join their SQL outputs. duckhts brings `read_gff`
(a queryable attribute `MAP`, tabix region scans), `read_vcf`/`read_bcf`, etc.;
duckvep brings the consequence engine. The block below is executed live in this
README: gene metadata comes from duckhts, consequence and HGVS from duckvep.

```{sql}
INSTALL duckhts FROM community;
LOAD duckhts;
WITH csq AS (
  SELECT c.gene_id, c.consequence, c.hgvsc
  FROM UNNEST(vep_consequence('17', 43124090, 'A', 'G')) AS u(c)
  WHERE c.canonical
), genes AS (
  SELECT attributes_map['gene_id'] AS gene_id, attributes_map['Name'] AS gene_name
  FROM read_gff('test/data/test.gff3.gz', region := '17:43044295-43170245',
                attributes_map := true)
  WHERE feature = 'gene'
)
SELECT g.gene_name, csq.consequence, csq.hgvsc
FROM csq JOIN genes g USING (gene_id);
```

A FASTA-backed, tabix-region version (adding `hgvsp`) is in
[`scripts/duckhts_integration_demo.R`](scripts/duckhts_integration_demo.R).

## Correctness

The accuracy oracle is **Ensembl VEP**. Concordance is measured version-matched (same
Ensembl release for VEP code, VEP cache and the gene model), keyed to the original input
variant, and split by IMPACT × variant class so an aggregate percentage cannot hide where
the misses are (high-impact, clinically actionable sites). The full report, rendered
directly from recorded CSVs, is **[`correctness/correctness.md`](correctness/correctness.md)**
(`make correctness`).

The table below reports error rate **per 100K pairs** vs offline Ensembl VEP, duckvep and
fastVEP side by side, generated from `correctness/data/concordance_by_impact.csv`. SNVs have
low error rates at every impact tier; the remaining discordance is concentrated in indels and
MNVs at exon/splice boundaries.

```{r correctness-headline, echo=FALSE, message=FALSE, results="asis"}
d <- read.csv("correctness/data/concordance_by_impact.csv")
d$rate <- sprintf("%d/100K", round((d$pairs - d$agree) / d$pairs * 1e5))  # error per 100K pairs
w <- reshape(d[, c("impact", "class", "engine", "rate")],
             idvar = c("impact", "class"), timevar = "engine", direction = "wide")
names(w) <- sub("^rate\\.", "", names(w))
w <- w[order(factor(w$impact, c("HIGH", "MODERATE", "LOW", "MODIFIER")), w$class), ]
cat(knitr::kable(w, col.names = c("impact", "class", "duckvep /100K", "fastVEP /100K"),
                 row.names = FALSE), sep = "\n")
```

Error rate **per 100,000 (variant, transcript) pairs** vs Ensembl VEP
(version-matched), by impact × class, duckvep vs fastVEP — generated from
`correctness/data/concordance_by_impact.csv`.

Impact × class is too coarse, though: it hides *which* SO categories fail and
*where duckvep regresses vs fastVEP*. Splitting the discordances by what VEP calls →
what duckvep calls, flagged by whether fastVEP matches VEP (`regression` = duckvep
worse than upstream; `shared` = inherited engine gap) — top by pair count,
generated from `correctness/data/error_transitions.csv`:

```{r correctness-regressions, echo=FALSE, message=FALSE, results="asis"}
et <- read.csv("correctness/data/error_transitions.csv", check.names = FALSE, stringsAsFactors = FALSE)
et$type <- ifelse(as.logical(et$duckvep_specific_regression), "regression", "shared")
et <- et[order(-et$n), c("type", "impact", "vep_calls", "duckvep_calls", "n")]
cat(knitr::kable(head(et, 8), col.names = c("type", "impact", "VEP calls", "duckvep calls", "n"),
                 row.names = FALSE, format.args = list(big.mark = ",")), sep = "\n")
```

The open work splits into **duckvep-specific regressions** (mainly splice sub-term precedence
on boundary indels, where VEP 3'-shifts the allele before calling consequence) and **shared
engine gaps** inherited from the vendored engine (`frameshift&stop_gained`, `missense`→
`synonymous`). Full per-SO-term and transition tables are in
[`correctness/correctness.md`](correctness/correctness.md); the accuracy patches over the
vendored engine are in [`docs/PATCHES.md`](docs/PATCHES.md).

## Benchmarks

Wall-clock and peak memory for duckvep and the fastVEP CLI on the benchmark inputs,
generated from `benchmarks/data/timings.csv`:

```{r benchmarks-timings, echo=FALSE, message=FALSE, results="asis"}
t <- read.csv("benchmarks/data/timings.csv", check.names = FALSE, stringsAsFactors = FALSE)
cat(knitr::kable(t[, c("dataset", "variants", "tool", "wall_clock", "peak_rss_mb")],
                 col.names = c("dataset", "variants", "tool", "wall", "peak RSS (MB)"),
                 format.args = list(big.mark = ",")), sep = "\n")
```

Full throughput / footprint / HGVS-concordance tables, and the methodology, are in
**[`benchmarks/results.md`](benchmarks/results.md)** (`make benchmarks`). Setup and
inputs: [`scripts/fetch-data.R`](scripts/fetch-data.R) (pinned versions).

## Roadmap

The main accuracy work is to port VEP's `TranscriptVariationAllele` coordinate layer (the
cDNA/CDS/protein mapper, `codon()`/`peptide()`, and the start/stop/frameshift predicates). VEP
3'-shifts an indel before calling consequence; duckvep does not yet, which is the source of the
`r dv_spec` duckvep-specific boundary-indel divergences in the conformance report. Porting that layer
replaces the current windowed-peptide surrogate with VEP's own coordinate model and is what would
turn the differential conformance framework from a falsifier into a soundness argument. Secondary
items — multi-transcript/opposite-strand witnesses, the haplotype path's compound-block flush, and
a native-DuckDB transcript cache — are tracked in [docs/refinements.md](docs/refinements.md).

## License

GPL-2.0-or-later. duckvep vendors parts of [fastVEP](https://github.com/Huang-lab/fastVEP)
(Apache-2.0, compatible with GPL ≥ 2) — see [`vendor/NOTICE.md`](vendor/NOTICE.md)
and [`docs/PATCHES.md`](docs/PATCHES.md) for the vendored crates and our divergences.
