---
output: github_document
---

<!-- README.md is generated from README.Rmd — edit the .Rmd and run `make readme`.
     SQL blocks below are executed live against the built extension via duckknit. -->

```{r setup, include=FALSE}
options(duckknit.duckdb = normalizePath("scripts/duckdb-unsigned"))
knitr::knit_engines$set(sql = duckknit::eng_duckdb)
knitr::opts_chunk$set(mode = "markdown", session = "readme",
                      comment = "", results = "asis")
```

# duckvep — DuckDB-native VEP: SQL genomics readers, VEP UDFs, Parquet sources

A loadable DuckDB extension (Rust,
[`duckdb-rs`](https://github.com/duckdb/duckdb-rs)) that reads genomics formats
via [noodles](https://github.com/zaeleus/noodles), exposes the VEP consequence /
HGVS / ACMG engine as SQL functions, and treats annotation databases as plain
Parquet/DuckDB tables joined by the optimizer — instead of hand-rolled file
formats. See [docs/DESIGN.md](docs/DESIGN.md) for the design, [NEWS.md](NEWS.md) for the
changelog, and [docs/PATCHES.md](docs/PATCHES.md) for our accuracy patches.

> **Status (v0.3.0).** Implemented: `read_vcf`/`vcf_samples`; `vep_consequence`
> (scan-driven scalar, incl. an END-aware form for structural variants) and
> `vep_annotate`; HGVS g./c./p.; `normalize_variant` (canonical minimal key for
> valid cross-annotator joins); a columnar Parquet transcript cache **built
> directly from Ensembl MySQL dumps** (inheriting MANE / `cds_start_NF` /
> selenocysteine / regulatory flags). Accuracy patches make duckvep *more*
> Ensembl-VEP-concordant than fastVEP (see [docs/PATCHES.md](docs/PATCHES.md)); concordance is
> reported version-matched and **split by impact × class** (see
> [correctness/](correctness/correctness.md)). Next: close the high-impact
> indel/MNV engine gap; supplementary-annotation joins; chrM/PAR/sex-chromosome
> correctness. Composes with the **duckhts** community extension in one session.

## Build

```sh
make debug      # builds build/debug/duckvep.duckdb_extension (native)
make test       # runs the SQL test suite
```

The extension also builds to **WASM**, so the same readers run in DuckDB-WASM in
the browser with no server.

## Load it

```{sql}
LOAD 'build/debug/duckvep.duckdb_extension';
```

## `read_vcf` — VCF/BCF as a SQL table

One row per variant. `alt` and `filter` are lists; `end_pos` carries the variant
interval (`INFO/END` for SV/CNV, else `pos + len(ref) - 1`).

```{sql}
SELECT chrom, pos, end_pos, ref, alt, qual, filter
FROM read_vcf('test/data/sites.vcf')
LIMIT 5;
```

### Structural variants are first-class

Symbolic (`<DEL>`, `<CNV>`), breakend, and multiallelic alleles survive as list
elements, and `end_pos` makes interval filters work:

```{sql}
SELECT pos, end_pos, ref, alt, end_pos - pos AS span
FROM read_vcf('test/data/sv.vcf')
WHERE end_pos - pos > 100
ORDER BY pos;
```

### Multi-sample, phased genotypes

`gt` is the positional per-sample genotype list (phasing preserved as `0|1`):

```{sql}
SELECT pos, ref, alt, gt
FROM read_vcf('test/data/ms.vcf');
```

`vcf_samples()` gives the header sample order, so genotypes explode to tidy
per-sample rows on demand — no re-annotation, because annotation is site-wise:

```{sql}
SELECT v.pos, s.sample, g.gt
FROM read_vcf('test/data/ms.vcf') v,
     UNNEST(v.gt) WITH ORDINALITY AS g(gt, idx)
     JOIN vcf_samples('test/data/ms.vcf') s USING (idx)
ORDER BY v.pos, s.idx;
```

### Region filter

```{sql}
SELECT count(*) AS n
FROM read_vcf('test/data/sv.vcf', region := 'chr1:4000-13000');
```

## Variant effect prediction

`vep_load_cache('<gene model>', '<fasta>')` builds the consequence engine **once**
into the connection's state, so later `vep_consequence`/`vep_annotate` calls reuse
it (the result `loaded` is just a confirmation). The gene model is either a GFF3
or duckvep's **columnar Parquet cache**: a GFF3 is parsed *and* written to
`<gff3>.transcripts.parquet` for next time; pass that `.parquet` to skip the parse.
The FASTA (`''` = none) is needed for exact coding calls (synonymous vs missense).

```{sql}
-- first call parses test.gff3 and writes test.gff3.transcripts.parquet;
-- next time, load the cache directly: vep_load_cache('test.gff3.transcripts.parquet','')
SELECT vep_load_cache('test/data/test.gff3', '');
```

Then `vep_consequence(chrom, pos, ref, alt)` is a scan-driven scalar returning a
native `LIST<STRUCT>` — `UNNEST` it. Driven by DuckDB's scan, so variants come
from `read_vcf`, `read_parquet`, or any relation.

Each struct also carries HGVS notation — `hgvsg` (genomic), `hgvsc` (coding) and
`hgvsp` (protein, when a FASTA is loaded) — exact-matching fastVEP's HGVS on chr17
(HGVSc 19,828/19,828, HGVSp 9,449/9,449):

```{sql}
SELECT c.transcript_id, c.consequence, c.impact, c.hgvsg, c.hgvsc
FROM UNNEST(vep_consequence('17', 43124090, 'A', 'G')) AS u(c)
WHERE c.gene_symbol = 'BRCA1'
ORDER BY c.canonical DESC
LIMIT 4;
```

Or annotate a whole VCF in one call with `vep_annotate(vcf, gff3 := …)` — one row
per (variant, transcript, allele), at parity with Ensembl VEP / fastVEP (same
engine):

```{sql}
SELECT pos, ref, alt, gene_symbol, transcript_id, consequence, impact
FROM vep_annotate('test/data/sites.vcf', gff3 := 'test/data/test.gff3')
WHERE canonical AND impact <> 'MODIFIER'
ORDER BY pos LIMIT 5;
```

### Structural variants

Symbolic SVs (`<DEL>`, `<DUP>`, `<CNV>`, `<INV>`) and breakends are dispatched to
the SV consequence predictor over their full `INFO/END` span — `vep_annotate`
reads `END` from the VCF, and the scalar takes an END-aware 5-argument form
`vep_consequence(chrom, pos, end, ref, alt)` (drive it with `read_vcf`'s
`end_pos`). A `<DEL>` covering a transcript yields `transcript_ablation`, a
partial one `feature_truncation` + `splice_*` + `copy_number_decrease`, and a
`<DUP>` `transcript_amplification` — the Ensembl SV vocabulary:

```{sql}
SELECT c.consequence, c.impact
FROM UNNEST(vep_consequence('17', 43044295, 43125483, 'N', '<DEL>')) AS u(c)
WHERE c.transcript_id = 'ENST00000357654';
```

## Variants from any source

Because the variant table is just columns, anything DuckDB can read is a valid
variant provider — `read_vcf`, `read_parquet`, `read_csv`, an attached DB, or a
literal relation. The downstream UDFs consume the columns, not the format:

```{sql}
SELECT chrom, pos, ref, alt
FROM (VALUES ('chr1', 100, 'A', ['G']),
             ('chr2', 200, 'C', ['T', 'TA'])) AS t(chrom, pos, ref, alt);
```

## Composing with duckhts

duckvep composes with **[duckhts](https://github.com/sounkou-bioinfo/duckhts)**
(htslib-backed format readers, a DuckDB **community extension**) by just loading
both and writing a `JOIN` — no glue code, no shared format. duckhts brings
`read_gff` (a queryable attribute `MAP`, tabix region scans), `read_vcf`/
`read_bcf`, etc.; duckvep brings the VEP engine. They share nothing but the
DuckDB ABI and SQL. This block is **executed live** in this README — gene
metadata comes from duckhts, consequence + HGVS from duckvep:

```{sql}
INSTALL duckhts FROM community;
LOAD duckhts;
WITH csq AS (
  SELECT c.gene_id, c.consequence, c.hgvsc
  FROM UNNEST(vep_consequence('17', 43124090, 'A', 'G')) AS u(c)
  WHERE c.canonical
), genes AS (
  SELECT attributes_map['gene_id'] AS gene_id, attributes_map['Name'] AS gene_name
  FROM read_gff('test/data/test.gff3.gz', region := '17:43044295-43170245',
                attributes_map := true)
  WHERE feature = 'gene'
)
SELECT g.gene_name, csq.consequence, csq.hgvsc
FROM csq JOIN genes g USING (gene_id);
```

A FASTA-backed, tabix-region version (adding `hgvsp`) is in
[`scripts/duckhts_integration_demo.sh`](scripts/duckhts_integration_demo.sh).

## Correctness

Our accuracy oracle is **Ensembl VEP itself**. Concordance is measured
version-matched (same Ensembl release for VEP code, VEP cache and the gene model)
and is **always split by IMPACT × variant class** — an aggregate % hides where the
misses are (high-impact, clinically actionable sites). The full report, rendered
directly from recorded CSVs, is **[`correctness/correctness.md`](correctness/correctness.md)**
(`make correctness`).

Headline — error rate **per 100K pairs** vs offline Ensembl VEP, duckvep and
fastVEP side by side, **generated from `correctness/data/concordance_by_impact.csv`**
(not hand-written): SNVs are near-perfect at every impact tier; the open frontier
is high-impact indels/MNVs (shared with fastVEP — an engine gap, not a duckvep bug).

```{r correctness-headline, echo=FALSE, message=FALSE, results="asis"}
d <- read.csv("correctness/data/concordance_by_impact.csv")
d$rate <- sprintf("%d/100K", round((d$pairs - d$agree) / d$pairs * 1e5))  # error per 100K pairs
w <- reshape(d[, c("impact", "class", "engine", "rate")],
             idvar = c("impact", "class"), timevar = "engine", direction = "wide")
names(w) <- sub("^rate\\.", "", names(w))
w <- w[order(factor(w$impact, c("HIGH", "MODERATE", "LOW", "MODIFIER")), w$class), ]
cat(knitr::kable(w, col.names = c("impact", "class", "duckvep /100K", "fastVEP /100K"),
                 row.names = FALSE), sep = "\n")
```

Error rate **per 100,000 (variant, transcript) pairs** vs Ensembl VEP
(version-matched), by impact × class, duckvep vs fastVEP — generated from
`correctness/data/concordance_by_impact.csv`.

duckvep's accuracy patches over the vendored engine are in [`docs/PATCHES.md`](docs/PATCHES.md).

## Benchmarks

Wall-clock + peak memory, duckvep vs fastVEP CLI (same consequence engine; the
difference is data-engineering — caching, streaming, columnar), **generated from
`benchmarks/data/timings.csv`**:

```{r benchmarks-timings, echo=FALSE, message=FALSE, results="asis"}
t <- read.csv("benchmarks/data/timings.csv", check.names = FALSE, stringsAsFactors = FALSE)
cat(knitr::kable(t[, c("dataset", "variants", "tool", "wall_clock", "peak_rss_mb")],
                 col.names = c("dataset", "variants", "tool", "wall", "peak RSS (MB)"),
                 format.args = list(big.mark = ",")), sep = "\n")
```

Full throughput / footprint / HGVS-concordance tables, and the methodology, are in
**[`benchmarks/results.md`](benchmarks/results.md)** (`make benchmarks`). Setup and
inputs: [`scripts/fetch-data.sh`](scripts/fetch-data.sh) (pinned versions).

## License

GPL-2.0-or-later. duckvep vendors parts of [fastVEP](https://github.com/Huang-lab/fastVEP)
(Apache-2.0, compatible with GPL ≥ 2) — see [`vendor/NOTICE.md`](vendor/NOTICE.md)
and [`docs/PATCHES.md`](docs/PATCHES.md) for the vendored crates and our divergences.
