---
output: github_document
---

<!-- correctness.md is generated from correctness.Rmd — edit the .Rmd and run
     `make correctness`. The table is read **directly from the recorded CSV**
     (correctness/data/concordance_by_impact.csv) and rendered with knitr. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
csv <- function(p) read.csv(p, check.names = FALSE, stringsAsFactors = FALSE)
```

# duckvep correctness

Our accuracy oracle is **Ensembl VEP itself** (decades of curated knowledge), not
fastVEP. Every figure below is read **directly from a recorded CSV** under
`correctness/data/`, regenerated by `correctness/vep_concordance.py` (offline
Ensembl VEP) and the synthetic corpus generator `correctness/gen_hard_variants.R`.

**Method (no slop):** VEP code, VEP cache and the duckvep GFF3 are the **same
Ensembl release** (no version skew), variants are biallelic ClinVar, and
concordance is the fraction of `(variant, transcript)` pairs whose `&`-joined SO
consequence set is identical. **We never report a single aggregate** — an average
hides *where* the misses are. fastVEP's headline "99.99%" concealed highly
significant discordances concentrated in HIGH-impact sites. So every figure is
split by **IMPACT × variant class**, and HIGH-impact discordances are surfaced.

## Concordance vs Ensembl VEP — split by IMPACT × class

`discordant / pairs` (lower discordant is better). duckvep is the same engine as
fastVEP plus our accuracy patches, so the head-to-head is fair.

```{r}
d <- csv("correctness/data/concordance_by_impact.csv")
d$disc <- d$pairs - d$agree
d$cell <- sprintf("%d / %d", d$disc, d$pairs)
# Accessor so every figure in the prose is generated from the CSV, never typed.
gv <- function(eng, imp, cls) {
  r <- d[d$engine == eng & d$impact == imp & d$class == cls, ]
  if (nrow(r) == 0) "n/a" else sprintf("%d/%d", r$disc, r$pairs)
}
w <- reshape(d[, c("impact", "class", "engine", "cell")],
             idvar = c("impact", "class"), timevar = "engine", direction = "wide")
names(w) <- sub("^cell\\.", "", names(w))
w <- w[order(factor(w$impact, c("HIGH", "MODERATE", "LOW", "MODIFIER")), w$class), ]
knitr::kable(w, col.names = c("impact", "class", "duckvep (disc/pairs)", "fastvep (disc/pairs)"),
             row.names = FALSE)
```

**Reading it** (all figures inline-generated from the CSV above):

- **SNVs are near-perfect at every impact tier** — duckvep `discordant/pairs`
  HIGH `r gv("duckvep","HIGH","snv")`, MODERATE `r gv("duckvep","MODERATE","snv")`,
  LOW `r gv("duckvep","LOW","snv")`, MODIFIER `r gv("duckvep","MODIFIER","snv")` —
  and ahead of fastVEP (MODERATE `r gv("fastvep","MODERATE","snv")`, LOW
  `r gv("fastvep","LOW","snv")`, MODIFIER `r gv("fastvep","MODIFIER","snv")`),
  thanks to the incomplete-CDS / non-ATG-start patches (see `../PATCHES.md`).
- **Discordances concentrate in HIGH-impact indels and MNVs** — duckvep HIGH del
  `r gv("duckvep","HIGH","del")`, HIGH ins `r gv("duckvep","HIGH","ins")`, HIGH mnv
  `r gv("duckvep","HIGH","mnv")` — the clinically actionable sites, and exactly what
  an aggregate % would bury. This is the open correctness frontier.
- **fastVEP's indel `pairs` are small and not comparable** (HIGH del
  `r gv("fastvep","HIGH","del")` vs duckvep `r gv("duckvep","HIGH","del")`):
  fastVEP's tab output reports anchor-stripped alleles that don't join on the
  original VCF allele, so most fastVEP indel rows drop out. Only the **SNV** and
  **MNV** fastVEP head-to-heads (which join cleanly) are valid — MNV is a shared
  engine gap (HIGH mnv `r gv("duckvep","HIGH","mnv")` in both).

## Synthetic hard-variant corpus

Random ClinVar sampling almost never places an indel *exactly* on a splice
landmark, so it cannot measure the splice/indel fixes. `gen_hard_variants.R`
constructs a corpus of pathological variants at **exact** offsets computed from
the transcript model — deletions/insertions spanning the donor, acceptor, 5th
base, donor region, polypyrimidine tract and splice region, plus CDS frameshift/
inframe indels and in-codon MNVs — to stress-test those classes directly.

## Reproduce

```sh
# 1. build the Ensembl-sourced cache (server-free, from MySQL dumps)
correctness/cache-build/build-cache.sh homo_sapiens 116 38
# 2. impact-stratified concordance vs offline Ensembl VEP
correctness/vep_concordance.py data/giab/clinvar.vcf.gz \
  data/giab/GRCh38.116.gff3.gz data/giab/GRCh38.primary.fa 50000
# 3. synthetic hard-variant corpus
correctness/gen_hard_variants.R \
  data/cache/homo_sapiens.116.38/exons.parquet data/giab/chr17.fa \
  data/cache/homo_sapiens.116.38/transcripts.parquet 17 correctness/data/hard_corpus.vcf
# 4. render this report
make correctness
```

See [`../PATCHES.md`](../PATCHES.md) for the accuracy patches and
[`cache-build/README.md`](cache-build/README.md) for the cache builder.
