---
output: github_document
---

<!-- correctness.md is generated from correctness.Rmd — edit the .Rmd and run
     `make correctness`. The table is read **directly from the recorded CSV**
     (correctness/data/concordance_by_impact.csv) and rendered with knitr. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
csv <- function(p) read.csv(p, check.names = FALSE, stringsAsFactors = FALSE)
```

# duckvep correctness

Our accuracy oracle is **Ensembl VEP itself** (decades of curated knowledge), not
fastVEP. Every figure below is read **directly from a recorded CSV** under
`correctness/data/`, regenerated by `correctness/vep_concordance.R` (offline
Ensembl VEP) and the synthetic corpus generator `tools/gen-hard-variants` (Rust + noodles).

All three annotators are compared on the **canonical normalized variant key**
(`normalize_variant`, matching VEP's minimal representation), so the join is valid
for *every* variant class including indels.

**Method (no slop):** VEP code, VEP cache and the duckvep GFF3 are the **same
Ensembl release** (no version skew), variants are biallelic ClinVar, and
concordance is the fraction of `(variant, transcript)` pairs whose `&`-joined SO
consequence set is identical. **We never report a single aggregate** — an average
hides *where* the misses are. fastVEP's headline "99.99%" concealed highly
significant discordances concentrated in HIGH-impact sites. So every figure is
split by **IMPACT × variant class**, and HIGH-impact discordances are surfaced.

**Controlled comparison — the only free variable is the engine.** VEP is run with
`--gff` on the **same GFF3 + FASTA the engines use** (not its own cache), so the
transcript *set* and *model* are identical for all three tools (duckvep, fastVEP, VEP);
only the engine differs. (Ensembl 116's GFF uses the `ncRNA_gene` feature type, which
VEP's hard-coded include-list omits — so it would silently drop ~42k non-coding genes;
we rewrite it to `gene`, identically for all three.) This collapses the cache-vs-GFF
transcript-set difference from **7,011 → 26** pairs, so the residual is pure engine. This
is *VEP 116's engine on the controlled GFF model* — it deliberately excludes cache-only
semantics (mature-miRNA products, RNA edits, selenocysteine), which is why e.g. the
`mature_miRNA_variant` discordances disappear under control.

**Scale vs the fastVEP paper.** The fastVEP manuscript validates accuracy as
"100% concordance across 23 annotation fields on **2,340 shared transcript-allele
pairs**" — which is the **173-variant VEP example file, all SNVs** — against VEP 115.1.
Concordance on a few thousand SNVs is the *easy* regime. This harness instead runs
**~1.4M `(variant, transcript)` pairs** (50,000 biallelic ClinVar variants, indels and
MNVs included) and scores the **consequence** SO-term set for *both* engines against the
same controlled VEP 116. That is where the divergences a 2,340-SNV set never surfaces
appear — concentrated, for fastVEP, in HIGH-impact indels and MNVs (table below).

## Concordance vs Ensembl VEP — split by IMPACT × class

`discordant / pairs` (lower discordant is better). duckvep is the vendored fastVEP
engine plus accuracy patches, so the head-to-head is fair; both are keyed to the
original input variant against the same controlled VEP 116.

```{r}
d <- csv("correctness/data/concordance_by_impact.csv")
.ma <- csv("correctness/data/methodology_audit.csv"); mav <- setNames(.ma$value, .ma$metric)
k <- function(m) format(mav[[m]], big.mark = ",")   # headline divergence figures, read from the audit CSV
d$disc <- d$pairs - d$agree
d$per100k <- round(d$disc / d$pairs * 1e5)          # error rate per 100,000 pairs
d$cell <- sprintf("%d (%d/100K)", d$disc, d$per100k)  # raw count + normalized rate
# Accessors so every figure in the prose is generated from the CSV, never typed.
gv <- function(eng, imp, cls) {
  r <- d[d$engine == eng & d$impact == imp & d$class == cls, ]
  if (nrow(r) == 0) "n/a" else sprintf("%d/%d", r$disc, r$pairs)
}
gr <- function(eng, imp, cls) {  # per-100K error rate
  r <- d[d$engine == eng & d$impact == imp & d$class == cls, ]
  if (nrow(r) == 0) "n/a" else sprintf("%d/100K", r$per100k)
}
# pairs now match across engines (valid join), so show it once.
w <- reshape(d[, c("impact", "class", "pairs", "engine", "cell")],
             idvar = c("impact", "class", "pairs"), timevar = "engine", direction = "wide")
names(w) <- sub("^cell\\.", "", names(w))
w <- w[order(factor(w$impact, c("HIGH", "MODERATE", "LOW", "MODIFIER")), w$class), ]
knitr::kable(w, col.names = c("impact", "class", "pairs", "duckvep err (rate)", "fastvep err (rate)"),
             row.names = FALSE)
```

Error rate is **per 100,000 (variant, transcript) pairs** — comparable across
classes despite very different pair counts.

**Reading it** (all figures inline-generated from the CSV above):

- **SNVs have low error rates at every impact tier** — duckvep error rate HIGH
  `r gr("duckvep","HIGH","snv")`, MODERATE `r gr("duckvep","MODERATE","snv")`,
  LOW `r gr("duckvep","LOW","snv")`, MODIFIER `r gr("duckvep","MODIFIER","snv")`,
  and below fastVEP (MODERATE `r gr("fastvep","MODERATE","snv")` vs duckvep
  `r gr("duckvep","MODERATE","snv")`), thanks to the incomplete-CDS / non-ATG-start
  patches (see `../docs/PATCHES.md`).
- **Indels and MNVs are the residual frontier**, concentrated at HIGH impact —
  duckvep error rate HIGH del `r gr("duckvep","HIGH","del")`, HIGH ins
  `r gr("duckvep","HIGH","ins")`, HIGH mnv `r gr("duckvep","HIGH","mnv")` (vs SNV
  `r gr("duckvep","HIGH","snv")`). These are the clinically actionable sites an
  aggregate % would bury.
- **The indel/MNV frontier is partly duckvep-specific.** Engines are compared keyed to
  the original input variant (not each engine's normalized form — VEP and duckvep 3'-shift
  indels differently, so a normalized-key join compares mismatched pairs and hides real
  regressions; that earlier join was a measurement artifact). On this keying, duckvep
  diverges on `r k("duckvep_discordant_on_shared")` shared pairs +
  `r format(mav[["duckvep_only_pairs_emission"]] + mav[["duckvep_missing_pairs_emission"]], big.mark=",")`
  emission = `r k("duckvep_total_divergence")` total, of which **`r k("duckvep_specific_on_shared")`
  are duckvep-specific** (fastVEP matches VEP), vs fastVEP's `r k("fastvep_total_divergence")`. Those
  are boundary indels where VEP 3'-shifts the allele before calling consequence; the rest are gaps
  shared with fastVEP.

## Convergence toward conformance

Each `correctness/vep_concordance.R` run appends a timestamped row of the headline divergence
counts to `correctness/data/conformance_history.csv`, so progress is tracked over time. The plot
decomposes duckvep's **total** divergence from VEP 116 into the part that is genuinely ours to fix
(**duckvep-specific** — fastVEP gets these right), the part **shared with fastVEP** (an inherited
engine gap), and **emission** differences (extra/missing pairs). The goal is to drive the whole
stack — and especially the duckvep-specific band — to zero.

```{r convergence, fig.width=8, fig.height=4.4, dpi=130}
h <- csv("correctness/data/conformance_history.csv")
h <- h[h$methodology == "original-identity-keyed", ]      # only comparable keyings on one axis
h$timestamp <- as.POSIXct(h$timestamp, format = "%Y-%m-%dT%H:%M:%S")
h <- h[order(h$timestamp), ]
# duckvep_total = duckvep-specific + shared-with-fastVEP + emission (the three bands sum to total)
comp <- rbind(
  data.frame(t = h$timestamp, band = "duckvep-specific (ours; fastVEP is right)", value = h$duckvep_specific),
  data.frame(t = h$timestamp, band = "shared with fastVEP (inherited engine gap)", value = h$duckvep_discordant_shared - h$duckvep_specific),
  data.frame(t = h$timestamp, band = "emission (extra / missing pairs)", value = h$duckvep_emission))
# stack order: duckvep-specific at the bottom (anchored at 0, easiest to read its shrinkage)
comp$band <- factor(comp$band, levels = c("emission (extra / missing pairs)",
  "shared with fastVEP (inherited engine gap)", "duckvep-specific (ours; fastVEP is right)"))
library(ggplot2)
ggplot(comp, aes(t, value, fill = band)) +
  geom_area(position = "stack", alpha = 0.92) +
  geom_point(data = h, aes(timestamp, duckvep_total), inherit.aes = FALSE, size = 1.6) +
  geom_text(data = h, aes(timestamp, duckvep_total, label = duckvep_total), inherit.aes = FALSE,
            vjust = -0.8, size = 3.4) +
  scale_fill_manual(values = c("duckvep-specific (ours; fastVEP is right)" = "#d62728",
                               "shared with fastVEP (inherited engine gap)" = "#9e9e9e",
                               "emission (extra / missing pairs)" = "#aec7e8"),
                    breaks = c("duckvep-specific (ours; fastVEP is right)",
                               "shared with fastVEP (inherited engine gap)",
                               "emission (extra / missing pairs)")) +
  labs(x = NULL, y = "duckvep divergences vs VEP 116", fill = NULL,
       title = "Convergence toward Ensembl VEP conformance",
       subtitle = sprintf("total %d → %d; duckvep-specific %d → %d (fastVEP control: %d)",
                          h$duckvep_total[1], tail(h$duckvep_total, 1),
                          h$duckvep_specific[1], tail(h$duckvep_specific, 1), tail(h$fastvep_total, 1))) +
  theme_minimal(base_size = 12) + theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 2))
```

The point labels are duckvep's total divergence; the red band at the base is the duckvep-specific
tail (the real target). The curve fills in as the engine is iterated. Raw history:
`conformance_history.csv`.

## Where the misses are — per consequence (SO term)

Impact × class is still too coarse: it buckets every "HIGH del" together while the
real culprits are **specific consequence categories** (splice acceptor/donor,
polypyrimidine tract, frameshift). This table explodes VEP's `&`-joined call and
attributes each discordant pair to the SO terms VEP assigned — the finest
"where is it actually failing" view. **Both engines are shown** (top HIGH terms by
duckvep discordant count) so the split between duckvep-specific regressions and gaps
shared with fastVEP is visible; generated from `correctness/data/discordance_by_consequence.csv`:

```{r echo=FALSE, message=FALSE}
b <- csv("correctness/data/discordance_by_consequence.csv")
dv <- b[b$engine == "duckvep", c("so_term", "impact", "pairs", "discordant", "per100k")]
fv <- b[b$engine == "fastvep", c("so_term", "impact", "discordant", "per100k")]
names(dv)[4:5] <- c("dv_disc", "dv_rate"); names(fv)[3:4] <- c("fv_disc", "fv_rate")
m <- merge(dv, fv, by = c("so_term", "impact"), all.x = TRUE)
m <- m[m$dv_disc > 0 & m$impact == "HIGH", ]
m <- m[order(-m$dv_disc), ]
m$duckvep <- sprintf("%d (%d/100K)", m$dv_disc, m$dv_rate)
m$fastvep <- ifelse(is.na(m$fv_disc), "n/a", sprintf("%d (%d/100K)", m$fv_disc, m$fv_rate))
# count terms where duckvep is strictly better / never worse than fastVEP (inline prose).
cmp <- m[!is.na(m$fv_disc), ]
n_better <- sum(cmp$dv_disc < cmp$fv_disc); n_worse <- sum(cmp$dv_disc > cmp$fv_disc)
knitr::kable(head(m[, c("so_term", "impact", "pairs", "duckvep", "fastvep")], 18),
             col.names = c("SO term (VEP)", "impact", "pairs", "duckvep disc (rate)", "fastvep disc (rate)"),
             row.names = FALSE, format.args = list(big.mark = ","))
```

A pair is discordant when its **whole** `&`-joined SO set differs; this table
credits that to each term VEP called, so a 100% rate means *every* variant whose
VEP call includes that term has a different full set. Reading the two engine
columns: duckvep is **better than fastVEP on `r n_better` of these terms and worse
on `r n_worse`** — i.e. duckvep is never worse than the upstream engine and ahead
on the frameshift / NMD / splice-donor fronts, while both still trail Ensembl. The
remaining frontier is **shared engine behaviour**: coding indels at exon/intron
boundaries where VEP co-calls `intron_variant` and the generic
`coding_sequence_variant`, frameshifts that introduce a premature stop
(`stop_gained`), and MNV codon edges (`missense`↔`synonymous`). **An impact-only
view hid exactly which categories to fix** — this is the precise frontier, and it
is the same for duckvep and fastVEP.

## Error transitions — what VEP calls vs what duckvep calls

The exact discordances: VEP's call → duckvep's call. The `type` column flags
whether fastVEP matches VEP — `regression` = **duckvep is worse than the upstream
engine** (something our patches broke), `shared` = both engines differ from VEP (an
inherited engine gap). Generated from `correctness/data/error_transitions.csv`.

```{r echo=FALSE, message=FALSE}
et <- csv("correctness/data/error_transitions.csv")
et$type <- ifelse(as.logical(et$duckvep_specific_regression), "regression", "shared")
et <- et[order(-et$n), c("type", "impact", "vep_calls", "duckvep_calls", "n")]
reg_pairs <- sum(et$n[et$type == "regression"])
knitr::kable(head(et, 16), col.names = c("type", "impact", "VEP calls", "duckvep calls", "n"),
             row.names = FALSE, format.args = list(big.mark = ","))
```

### Split by SO term (fair to fastVEP)

The table above keys on the **whole** consequence string, so a one-term difference is
hidden inside shared co-occurring terms (`NMD_transcript_variant`, 5′/3′ UTR). This
strips each discordance to the **symmetric difference** — only the SO terms VEP and
duckvep disagree on — and judges fastVEP **term-by-term**: a `regression` here means
fastVEP is correct on *exactly* the terms duckvep gets wrong (has every term duckvep
missed, adds none duckvep added), so a shared `c1&c2` gap is never mis-blamed on duckvep
and a genuine single-term fastVEP win is not masked by the rest of its string. Generated
from `correctness/data/so_term_transitions.csv`.

```{r echo=FALSE, message=FALSE}
st <- csv("correctness/data/so_term_transitions.csv")
st$type <- ifelse(as.logical(st$duckvep_specific_regression), "regression", "shared")
st <- st[order(-st$n), c("type", "impact", "vep_terms", "duckvep_terms", "n")]
st_reg <- sum(st$n[st$type == "regression"])
knitr::kable(head(st, 16), col.names = c("type", "impact", "VEP term(s)", "duckvep term(s)", "n"),
             row.names = FALSE, format.args = list(big.mark = ","))
```

At the SO-term level, **`r format(st_reg, big.mark=",")` pairs** are duckvep-specific
(fastVEP right on the exact differing term) — these are the only places duckvep is worse
than upstream on a term; everything else is a shared engine gap or a term fastVEP also
misses.

**duckvep-specific regressions (whole-string): `r format(reg_pairs, big.mark=",")` pairs** (a
`regression` is a discordance where fastVEP *matches* VEP, so our patches — not the
inherited engine — are responsible). The interval-aware splice rewrite originally
introduced ~1,900 such pairs; root-causing them against Ensembl
`BaseTranscriptVariationAllele.pm` closed them in three steps: the polypyrimidine
`(min,max)` swap and `_intron_overlap` insertion cases (283 insertion under-calls),
the codon-table-aware non-ATG `start_lost` (mito initiators), and a harness join-key
fix (deletions share `alt='-'`, so the key must include `ref`). What remains are all
**shared engine gaps** (both duckvep *and* fastVEP differ from VEP): a frameshift
introducing a premature stop (`frameshift_variant&stop_gained` → `frameshift_variant`)
and MNV codon edges — real engine accuracy work vs Ensembl, not something our patches
broke.

## Auditing the fastVEP paper's "100% concordance" claim

The fastVEP manuscript reports *"100% concordance across 23 annotation fields on 2,340
shared (allele, transcript) pairs"* vs Ensembl VEP 115.1. We reproduced its exact
protocol — the VEP example dataset (`examples/homo_sapiens_GRCh38.vcf`) with
`--symbol --canonical --hgvs` — and audited two words doing the heavy lifting.

**"example dataset" → all SNVs.** The 173 example variants are **173/173 SNVs, zero
indels, zero MNVs** (71% are synonymous+missense). duckvep — the *same* engine plus our
patches — is **also 100%** on it (5,044/5,044 shared pairs). 100% concordance on a few
thousand SNVs is the easy regime; it says nothing about the classes where the engine
diverges.

**"shared pairs" → the transcript-set difference was a cache-vs-GFF artifact, now
controlled away.** With VEP run on its **cache**, it emitted ~7,000 fewer
`(variant,transcript)` pairs than the GFF-driven engines (the paper calls these "lncRNA"
— in fact they are *mostly protein-coding*, and VEP also *missed* ~400 the engines emit,
contradicting "no missing transcripts"). Once VEP reads the **same GFF model** (`--gff`,
above), that gap collapses to the handful below, and the residual is pure engine.
Numbers read from `correctness/data/methodology_audit.csv`:

A shared-pair inner join could *hide* a tool failing to emit a transcript (it would
become a non-shared row, not a discordance), so we count **every** kind of divergence as
first-class: consequence disagreement on shared pairs, **plus** pairs each tool emits
that the oracle does not (extra), **plus** pairs the oracle emits that the tool misses.
Read from `correctness/data/methodology_audit.csv`:

```{r echo=FALSE, message=FALSE}
ma <- csv("correctness/data/methodology_audit.csv")
mv <- function(k) as.integer(ma$value[ma$metric == k])
audit <- data.frame(
  divergence = c("consequence disagreement (shared pairs)",
                 "extra pairs the tool emits, VEP does not",
                 "pairs VEP emits, the tool misses",
                 "TOTAL divergence vs VEP 116"),
  duckvep = c(mv("duckvep_discordant_on_shared"), mv("duckvep_only_pairs_emission"),
              mv("duckvep_missing_pairs_emission"), mv("duckvep_total_divergence")),
  fastVEP = c(mv("fastvep_discordant_on_shared"), mv("fastvep_only_pairs_emission"),
              mv("fastvep_missing_pairs_emission"), mv("fastvep_total_divergence")))
knitr::kable(audit, col.names = c("divergence type (controlled, 50K ClinVar)", "duckvep", "fastVEP"),
             row.names = FALSE, format.args = list(big.mark = ","))
```

With the gene model controlled, the emission gap is tiny in both directions, so the
comparison is genuinely pure-engine — and counting *all* divergence (consequence + extra
+ missing), **duckvep totals `r format(mv("duckvep_total_divergence"), big.mark=",")`
vs fastVEP `r format(mv("fastvep_total_divergence"), big.mark=",")`** of ~1.43M pairs. The
duckvep residual is dominated by consequence-engine edges (start-codon straddle, boundary
insertion); the 27 extras are 5′/3′-distance-window boundary cases. The fastVEP paper's
100% held only because its validation set is 173 SNVs.

## Synthetic hard-variant corpus

Random ClinVar sampling almost never places an indel *exactly* on a splice
landmark, so it cannot measure the splice/indel fixes. `tools/gen-hard-variants`
(a Rust binary using noodles + the parquet cache — no R/bash/Python) constructs a
corpus of pathological variants at **exact** offsets computed from the transcript
model — deletions/insertions spanning the donor, acceptor, 5th base, donor region,
polypyrimidine tract and splice region, plus CDS frameshift/inframe indels and
in-codon MNVs — to stress-test those classes directly.

## Reproduce

```sh
# 0. fetch all inputs with pinned provenance (Ensembl 116, ClinVar 2026-06-06,
#    VEP cache 116, GIAB) -> data/ ; see scripts/fetch-data.R for exact sources
scripts/fetch-data.R
# 1. build the Ensembl-sourced cache (server-free, from MySQL dumps)
correctness/cache-build/build-cache.R homo_sapiens 116 38
# 2. impact-stratified concordance vs offline Ensembl VEP
correctness/vep_concordance.R data/giab/clinvar.vcf.gz \
  data/giab/GRCh38.116.gff3.gz data/giab/GRCh38.primary.fa 50000
# 3. synthetic hard-variant corpus (Rust + noodles)
cargo build -p duckvep-tools --release
./target/release/gen-hard-variants \
  data/cache/homo_sapiens.116.38/exons.parquet \
  data/cache/homo_sapiens.116.38/transcripts.parquet \
  data/giab/chr17.fa 17 correctness/data/hard_corpus.vcf
# 4. render this report
make correctness
```

See [`../docs/PATCHES.md`](../docs/PATCHES.md) for the accuracy patches and
[`cache-build/README.md`](cache-build/README.md) for the cache builder.
