---
output: github_document
---

<!-- correctness.md is generated from correctness.Rmd — edit the .Rmd and run
     `make correctness`. The table is read **directly from the recorded CSV**
     (correctness/data/concordance_by_impact.csv) and rendered with knitr. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
csv <- function(p) read.csv(p, check.names = FALSE, stringsAsFactors = FALSE)
```

# duckvep correctness

Our accuracy oracle is **Ensembl VEP itself** (decades of curated knowledge), not
fastVEP. Every figure below is read **directly from a recorded CSV** under
`correctness/data/`, regenerated by `correctness/vep_concordance.py` (offline
Ensembl VEP) and the synthetic corpus generator `tools/gen-hard-variants` (Rust + noodles).

All three annotators are compared on the **canonical normalized variant key**
(`normalize_variant`, matching VEP's minimal representation), so the join is valid
for *every* variant class including indels.

**Method (no slop):** VEP code, VEP cache and the duckvep GFF3 are the **same
Ensembl release** (no version skew), variants are biallelic ClinVar, and
concordance is the fraction of `(variant, transcript)` pairs whose `&`-joined SO
consequence set is identical. **We never report a single aggregate** — an average
hides *where* the misses are. fastVEP's headline "99.99%" concealed highly
significant discordances concentrated in HIGH-impact sites. So every figure is
split by **IMPACT × variant class**, and HIGH-impact discordances are surfaced.

## Concordance vs Ensembl VEP — split by IMPACT × class

`discordant / pairs` (lower discordant is better). duckvep is the same engine as
fastVEP plus our accuracy patches, so the head-to-head is fair.

```{r}
d <- csv("correctness/data/concordance_by_impact.csv")
d$disc <- d$pairs - d$agree
d$per100k <- round(d$disc / d$pairs * 1e5)          # error rate per 100,000 pairs
d$cell <- sprintf("%d (%d/100K)", d$disc, d$per100k)  # raw count + normalized rate
# Accessors so every figure in the prose is generated from the CSV, never typed.
gv <- function(eng, imp, cls) {
  r <- d[d$engine == eng & d$impact == imp & d$class == cls, ]
  if (nrow(r) == 0) "n/a" else sprintf("%d/%d", r$disc, r$pairs)
}
gr <- function(eng, imp, cls) {  # per-100K error rate
  r <- d[d$engine == eng & d$impact == imp & d$class == cls, ]
  if (nrow(r) == 0) "n/a" else sprintf("%d/100K", r$per100k)
}
# pairs now match across engines (valid join), so show it once.
w <- reshape(d[, c("impact", "class", "pairs", "engine", "cell")],
             idvar = c("impact", "class", "pairs"), timevar = "engine", direction = "wide")
names(w) <- sub("^cell\\.", "", names(w))
w <- w[order(factor(w$impact, c("HIGH", "MODERATE", "LOW", "MODIFIER")), w$class), ]
knitr::kable(w, col.names = c("impact", "class", "pairs", "duckvep err (rate)", "fastvep err (rate)"),
             row.names = FALSE)
```

Error rate is **per 100,000 (variant, transcript) pairs** — comparable across
classes despite very different pair counts.

**Reading it** (all figures inline-generated from the CSV above):

- **SNVs are near-perfect at every impact tier** — duckvep error rate HIGH
  `r gr("duckvep","HIGH","snv")`, MODERATE `r gr("duckvep","MODERATE","snv")`,
  LOW `r gr("duckvep","LOW","snv")`, MODIFIER `r gr("duckvep","MODIFIER","snv")` —
  and ahead of fastVEP (MODERATE `r gr("fastvep","MODERATE","snv")` vs duckvep
  `r gr("duckvep","MODERATE","snv")`), thanks to the incomplete-CDS / non-ATG-start
  patches (see `../docs/PATCHES.md`).
- **Indels and MNVs are the residual frontier**, concentrated at HIGH impact —
  duckvep error rate HIGH del `r gr("duckvep","HIGH","del")`, HIGH ins
  `r gr("duckvep","HIGH","ins")`, HIGH mnv `r gr("duckvep","HIGH","mnv")` (vs SNV
  `r gr("duckvep","HIGH","snv")`). These are the clinically actionable sites an
  aggregate % would bury.
- **duckvep and fastVEP are now validly comparable on indels** — normalizing both
  sides makes the indel `pairs` match across engines (HIGH del fastVEP
  `r gv("fastvep","HIGH","del")` vs duckvep `r gv("duckvep","HIGH","del")`; before
  normalization fastVEP joined only a small unrepresentative subset). The two are
  nearly identical, so the indel/MNV frontier is **shared engine behavior** vs
  Ensembl VEP — a real engine gap to close, not duckvep-specific and not a
  measurement artifact.

## Where the misses are — per consequence (SO term)

Impact × class is still too coarse: it buckets every "HIGH del" together while the
real culprits are **specific consequence categories** (splice acceptor/donor,
polypyrimidine tract, frameshift). This table explodes VEP's `&`-joined call and
attributes each discordant pair to the SO terms VEP assigned — the finest
"where is it actually failing" view (duckvep, top by discordant count, generated
from `correctness/data/discordance_by_consequence.csv`):

```{r echo=FALSE, message=FALSE}
b <- csv("correctness/data/discordance_by_consequence.csv")
b <- b[b$engine == "duckvep" & b$discordant > 0, ]
b <- b[order(-b$discordant), c("so_term", "impact", "pairs", "discordant", "per100k")]
b$per100k <- sprintf("%d/100K", b$per100k)
knitr::kable(head(b, 18),
             col.names = c("SO term (VEP)", "impact", "pairs", "discordant", "rate"),
             row.names = FALSE, format.args = list(big.mark = ","))
```

A pair is discordant when its **whole** `&`-joined SO set differs; this table
credits that to each term VEP called, so a 100% rate means *every* variant whose
VEP call includes that term has a different full set in duckvep. The
splice-region sub-terms (`splice_donor_5th_base`, `splice_donor_region`,
`splice_polypyrimidine_tract`, co-occurring `intron_variant`) and
`coding_sequence_variant`/`stop_retained_variant` dominate — i.e. the boundaries
are right (interval predicates) but the splice-region term *combinations* and the
incomplete-CDS / stop edges still diverge from VEP. **An impact-only view hid
exactly which categories to fix** — this is the precise frontier.

## Error transitions — what VEP calls vs what duckvep calls

The exact discordances: VEP's call → duckvep's call. The `type` column flags
whether fastVEP matches VEP — `regression` = **duckvep is worse than the upstream
engine** (something our patches broke), `shared` = both engines differ from VEP (an
inherited engine gap). Generated from `correctness/data/error_transitions.csv`.

```{r echo=FALSE, message=FALSE}
et <- csv("correctness/data/error_transitions.csv")
et$type <- ifelse(as.logical(et$duckvep_specific_regression), "regression", "shared")
et <- et[order(-et$n), c("type", "impact", "vep_calls", "duckvep_calls", "n")]
reg_pairs <- sum(et$n[et$type == "regression"])
knitr::kable(head(et, 16), col.names = c("type", "impact", "VEP calls", "duckvep calls", "n"),
             row.names = FALSE, format.args = list(big.mark = ","))
```

**duckvep-specific regressions: `r format(reg_pairs, big.mark=",")` pairs** (a
`regression` is a discordance where fastVEP *matches* VEP, so our patches — not the
inherited engine — are responsible). The interval-aware splice rewrite originally
introduced ~1,900 such pairs; root-causing them against Ensembl
`BaseTranscriptVariationAllele.pm` closed them in three steps: the polypyrimidine
`(min,max)` swap and `_intron_overlap` insertion cases (283 insertion under-calls),
the codon-table-aware non-ATG `start_lost` (mito initiators), and a harness join-key
fix (deletions share `alt='-'`, so the key must include `ref`). What remains are all
**shared engine gaps** (both duckvep *and* fastVEP differ from VEP): a frameshift
introducing a premature stop (`frameshift_variant&stop_gained` → `frameshift_variant`)
and MNV codon edges — real engine accuracy work vs Ensembl, not something our patches
broke.

## Synthetic hard-variant corpus

Random ClinVar sampling almost never places an indel *exactly* on a splice
landmark, so it cannot measure the splice/indel fixes. `tools/gen-hard-variants`
(a Rust binary using noodles + the parquet cache — no R/bash/Python) constructs a
corpus of pathological variants at **exact** offsets computed from the transcript
model — deletions/insertions spanning the donor, acceptor, 5th base, donor region,
polypyrimidine tract and splice region, plus CDS frameshift/inframe indels and
in-codon MNVs — to stress-test those classes directly.

## Reproduce

```sh
# 0. fetch all inputs with pinned provenance (Ensembl 116, ClinVar 2026-06-06,
#    VEP cache 116, GIAB) -> data/ ; see scripts/fetch-data.sh for exact sources
scripts/fetch-data.sh
# 1. build the Ensembl-sourced cache (server-free, from MySQL dumps)
correctness/cache-build/build-cache.sh homo_sapiens 116 38
# 2. impact-stratified concordance vs offline Ensembl VEP
correctness/vep_concordance.py data/giab/clinvar.vcf.gz \
  data/giab/GRCh38.116.gff3.gz data/giab/GRCh38.primary.fa 50000
# 3. synthetic hard-variant corpus (Rust + noodles)
cargo build -p duckvep-tools --release
./target/release/gen-hard-variants \
  data/cache/homo_sapiens.116.38/exons.parquet \
  data/cache/homo_sapiens.116.38/transcripts.parquet \
  data/giab/chr17.fa 17 correctness/data/hard_corpus.vcf
# 4. render this report
make correctness
```

See [`../docs/PATCHES.md`](../docs/PATCHES.md) for the accuracy patches and
[`cache-build/README.md`](cache-build/README.md) for the cache builder.
