---
output: github_document
---

<!-- results.md is generated from results.Rmd — edit the .Rmd and run `make benchmarks`.
     Tables are read **directly from the recorded CSVs** and rendered with knitr. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
csv <- function(p) read.csv(p, check.names = FALSE, stringsAsFactors = FALSE)
kbl <- function(df, ...) knitr::kable(df, ...)
```

# duckvep benchmarks

R Markdown report — every table below is read **directly from a recorded CSV**
under `benchmarks/data/` (regenerate the data with `scripts/benchmark.sh` /
`correctness/vep_concordance.py`, then `make benchmarks`). duckvep wraps the same
consequence engine as fastVEP, so head-to-heads are fair; the differences are
data-engineering (caching, streaming, columnar), not the science. **Setup:**
Ensembl GFF3 gene models; release builds (`opt-level=3`, `lto=fat`); steady state
(warm transcript cache); single run each.

> Scope: duckvep currently emits **consequence + amino-acids/codons**; HGVS
> strings and supplementary annotations are not yet wired, so throughput is
> "consequence", not "full annotation".

## Throughput (consequence)

```{r}
t <- csv("benchmarks/data/throughput.csv")
t[["variants/s"]] <- formatC(round(t$variants / t$time_s), format = "d", big.mark = ",")
t$variants <- formatC(t$variants, format = "d", big.mark = ",")
t$transcripts <- formatC(t$transcripts, format = "d", big.mark = ",")
t$time <- sprintf("%.1fs", t$time_s)
kbl(t[, c("organism", "assembly", "transcripts", "variants", "source", "time", "variants/s")])
```

Add organisms by running `scripts/benchmark.sh "<organism>" <assembly> <gff3>
<vcf> [fasta]` (yeast / fly / arabidopsis / mouse / human).

## Head-to-head vs fastVEP (full output, same gene model)

```{r}
kbl(csv("benchmarks/data/timings.csv"))
```

duckvep's edge is **caching** the parsed transcript model (warm); a *cold* run
(parsing the GFF3) is ~parity with fastVEP — same engine. Streaming `read_vcf`
keeps memory bounded (full GIAB 2.0 GB vs 7.3 GB eager). Offline Ensembl-VEP
timing is pending its cache install; concordance vs the live VEP is below.

## Footprint

```{r}
kbl(csv("benchmarks/data/footprint.csv"))
```

## Memory & parallelism

- **Streaming `read_vcf`** bounds memory to one ~2048-row chunk: full GIAB
  **7.3 GB → 2.0 GB**.
- **Columnar Parquet transcript cache** (zstd ~45 MB, `read_parquet`-able);
  warm load ~1.1 s vs ~5.7 s GFF3 parse.
- **`Arc<str>`** shares the repeated categorical fields across output rows.
- **Parallel scan:** the consequence scalar is thread-safe (lock-free `ArcSwap`
  cache), so over `read_parquet` DuckDB runs it on multiple cores
  (**100% → 199% CPU**); `read_vcf` itself streams single-threaded.

## Ensembl-VEP concordance

Per-(variant, transcript) consequence agreement vs the **live Ensembl VEP** (REST
API), over sampled real ClinVar variants annotated with the reference FASTA (so
synonymous vs missense is exact). Both duckvep **and** fastVEP vs Ensembl VEP:

```{r}
kbl(csv("data/vep_dumps/concordance_log.csv"))
```

duckvep matches Ensembl VEP because it wraps the same engine; the table also
shows fastVEP itself vs Ensembl VEP. Dated annotation dumps accumulate under
`data/vep_dumps/<date>/annotations.parquet`.
